The Talking Heads Experiment

### 2015-04-23 chap7/chap7.tex
cleaned up after sanity check
fixed some more widows/orphans
hyphenation, punctuation, positioning, etc for chapters 1-6
hyphenations, footnotes, ...

But some drastic events happened in Wittgenstein's life which 
caused a radical shift in his philosophical orientation. 
One of them was that he became headmaster of a small
school in the Austrian mountains,
without losing his interest for philosophising about language and 
meaning. By working with children and by seeing concretely
how they engaged with language, Wittgenstein realised that the 
logicist approach did not address some basic questions one 
can ask about language, particularly how the semantics of a 
language might arise, or how people ever develop a shared 
communication system. Wittgenstein
saw clearly that a language imposes a certain view on the 
world, which gives each language its own strength as well as
its limitations. The logicist 
framework might be ideal as a tool for constructing a 
{\itshape post factum} formal description of the semantics of
an (ideal) language, just as a structuralist grammar identifies
the frozen state of its syntax, but
not for modeling how language achieves its communicative purpose, 
comes about, or evolves. 
\enlargethispage{1\baselineskip}
Wittgenstein proposed to study the use of language in terms of 
games. This notion captures that the most basic form of
language involves a social interaction between individuals
within a specific setting, which acts as a context restricting
the possible meanings. The interaction is played out 
along a conventionalised set of rules, as is chess or any
other game. The notion of a game captures
that a language interaction has a purpose, for example,
to identify an object in a context, to gather 
information, to transmit emotion, to try and
invoke an action by another person, etc. Words get 
their meaning as part of a language game. Language is a tool 
fully integrated in the rest of human social activity.\footnote{See in particular \cite{Wittgenstein:1953}.}

@@

\begin{table}
\begin{center}
\begin{tabular}{lllll}
\lsptoprule
{\itshape Word}&{\itshape Meaning}&{\itshape Translation} & {\bfshape a1}&{\bfshape a2} \\ \midrule
wawosido & [\textsc{hpos} 0.0–0.5] &left&0.4&0.4\\ 
meluri & [\textsc{hpos} 0.25–0.5] &medium left&0.1&0.0\\ 
gofubo & [\textsc{hpos} 0.5–1.0]& right&1.0&1.0\\ 
wiwigapo & [\textsc{vpos} 0.0–0.5] &left&0.1&0.0\\ 
fozumoba & [\textsc{area} 0.5–1.0]&large & 0.0&0.1\\ 
wefoto & [R 0.0–0.5]& low redness &0.2&0.2\\ 
togene & [R 0.5–1.0]& high redness &0.5&1.0\\ 
fumudanu & [G 0.0–0.5]& low greenness &0.2&0.2\\ 
puxedu & [G 0.5–1.0]& high greenness &0.4&0.4\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game50}First words after about a dozen games.}
\end{center}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{lllll}
\lsptoprule
{\itshape Word}&{\itshape Meaning}&{\itshape Translation} & {\bfshape a1}&{\bfshape a2} \\ \midrule
wawosido & [\textsc{hpos} 0.0–0.5] &left&0.7&0.5\\ 
meluri & [\textsc{hpos} 0.25–0.5] &medium left&0.4&0.4\\ 
gofubo & [\textsc{hpos} 0.5–1.0]& right&1.0&1.0\\ 
vokomutu & [\textsc{hpos} 0.5–0.75] &medium right&0.2&0.2\\ 
buwonipo & [\textsc{hpos} 0.75–0.875] &strongly right&0.1&0.0\\ 
wiwigapo & [\textsc{vpos} 0.0–0.5] &down&0.6&0.6\\ 
fozumoba & [\textsc{area} 0.5–1.0]&large & 0.2&0.4\\ 
wefoto & [R 0.0–0.5]& low redness &0.2&0.3\\ 
togene & [R 0.5–1.0]& high redness &1.0&1.0\\ 
fumudanu & [G 0.0–0.5]& low greenness &0.7&0.7\\ 
puxedu & [G 0.5–1.0]& high greenness &0.7&0.9\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:goubo}Population lexicon after 100 games.}
\end{center}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{lll}
\lsptoprule
{\itshape channel}& {\itshape obj-0} & {\itshape obj-1}\\ \midrule
\textsc{hpos} & 0.08 & 0.09\\ 
\textsc{vpos} & 0.21 & 0.10\\ 
\textsc{height} & 0.10 & 0.15\\ 
\textsc{width} & 0.42 & 0.11\\ 
\textsc{area} & 0.34 & 0.16\\ 
R & 0.33 & 0.31\\ 
G & 0.68 & 0.65\\ 
B & 0.52 & 0.50\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game126}Sensory data for game 126.}
\end{center}
\end{table}
Clearly width is the most salient channel for the speaker. 
It is therefore chosen and because the 
agents have already grown distinctions on this channel, 
discrimination succeeds and a new word can be
constructed and stored. 
\begin{verbatim}
Game 126 
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 2 objects: 
       object-0 object-1
  a2 chooses object-0 as the topic 
  a2 categorises the topic as [\textsc{width} 0.5–1.0]
  a2 creates a new word: vaviwumu
  a2 says: vaviwumu
  a1 does not know vaviwumu
  a1 says: vaviwumu?
  a2 points to object-0
  a1 categorises the topic as [\textsc{width} 0.5–1.0]
  a1 stores vaviwumu as [\textsc{width} 0.5–1.0]
\end{verbatim}
\figref{fig:plate-13} (bottom) contains another example
of segmented images which have stimulated conceptual
growth and hence an expansion of the lexicon. 
In this case, categorisations on the \textsc{height} channel are
relevant and a word developed for tall.

@@

\begin{table}
\begin{center}
\begin{tabular}{lllll}
\lsptoprule
{\itshape Word}&{\itshape Meaning}&{\itshape Translation} & {\bfshape a1}&{\bfshape a2} \\ \midrule
wawosido & [\textsc{hpos} 0.0–0.5] &left&0.9&0.4\\ 
gixepo & [\textsc{hpos} 0.0–0.25] & very left&0.4&0.4\\ 
wonuxa & [\textsc{hpos} 0.0–0.25] & very left&0.10&0.0\\ 
meluri & [\textsc{hpos} 0.25–0.5] &medium left&0.4&0.4\\ 
gofubo & [\textsc{hpos} 0.5–1.0]& right&1.0&1.0\\ 
vokomutu & [\textsc{hpos} 0.5–0.75] &medium right&0.4&0.4\\ 
buwonipo & [\textsc{hpos} 0.75–0.875] &strongly right&0.1&0.0\\ 
wiwigapo & [\textsc{vpos} 0.0–0.5] &down&0.5&0.4\\ 
putuwenu & [\textsc{vpos} 0.5–1.0]&up & 0.5&0.5\\ 
vaviwumu & [\textsc{width} 0.5–1.0]&wide & 0.5&0.5\\ 
pesidumu & [\textsc{area} 0.0–0.5]&small& 0.2&0.2\\ 
fozumoba & [\textsc{area} 0.5–1.0]&large & 1.0&1.0\\ 
wefoto & [R 0.0–0.5]& low redness &0.2&0.1\\ 
togene & [R 0.5–1.0]& high redness &0.9&1.0\\ 
fumudanu & [G 0.0–0.5]& low greenness &0.7&0.9\\ 
puxedu & [G 0.5–1.0]& high greenness &0.9&0.9\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:upper}Population lexicon after 200 games.}
\end{center}
\end{table}
Note that new words have come into the lexicon 
for up (`putuwenu'), down (`wiwigapo'), and 
wide (`vaviwumu'). The discrimination trees of 
the \textsc{width} and \textsc{height} channel have started to 
expand (\figref{discri200} bottom).

@@

Grounding language games in physical environments\is{perceptual anomalies}
makes it obviously harder
for the agents in a number of respects. First of 
all because perception and segmentation may differ, 
the salient characteristics of objects may not be the same
and as a result the hearer may guess another meaning for an 
unknown word, compared to 
the one used by the speaker. This happens for
example in the following game (game 128) which is based on 
the segmented images shown in \figref{fig:plate-16} (top). The
speaker's perception is shown to the left and the 
hearer's to the right. 
\begin{verbatim}
Game 128
  a1 is the speaker. a2 is the hearer. 
  a1 segments the context into 3 objects: 
       object-0 object-1 object-2
  a1 chooses object-0 as the topic 
  a1 categorises the topic as 
       [\textsc{hpos} 0.0–0.25] [\textsc{hpos} 0.0–0.125]
  a1 creates a new word: `bagaxe' for [\textsc{hpos} 0.0–0.25]
  a1 says: bagaxe
  a2 does not know bagaxe
  a2 says: bagaxe?
  a1 points to object-0
  a2 categorises the topic as [\textsc{area} 0.5–1.0]
  a2 stores bagaxe as [\textsc{area} 0.5–1.0]
\end{verbatim}

@@

Due to perceptual anomalies, a game may fail even though 
both agents already have the same meaning for the 
same word. This happens when this meaning yields different 
objects (or no objects at all) for the speaker and the hearer. 
As a consequence, the hearer adopts another meaning for 
the word which starts to compete with the one he already had. 
This happened in the following game (game 127)
based on the segmented images shown in \figref{fig:plate-16} (bottom). The speaker (left image) has identified
only two objects, but the hearer three. The 
speaker (left image) uses green as distinguishing 
characteristic which is indeed appropriate. but 
because the hearer's third object (the top rectangle) is 
also green, this distinction fails for him.  
Even though the hearer had already a well established
meaning for `puxedu', he associates a new meaning 
based on the \textsc{width}-channel with this word because this 
channel is now the most salient. 
\begin{verbatim}
Game 127
  a1 is the speaker. a2 is the hearer. 
  a1 segments the context into 2 objects: 
       s-object-0 s-object-1
  a1 chooses object-0 as the topic 
  a1 considers as salient G R \textsc{area} \textsc{hpos} 
  a1 categorises the topic as 
       [\textsc{hpos} 0.0–0.5] [\textsc{hpos} 0.0–0.125]
       [\textsc{area} 0.0–0.5] [R 0.0–0.5] [G 0.5–1.0]
  a1 has the words
       puxedu for [G 0.5–1.0] (1.0)
       wawosido for [\textsc{hpos} 0.0–0.5] (0.7)
       pesidimu for [\textsc{area} 0.0–0.5] (0.2)
       wefoto for [R 0.0–0.5] (0.2)
       buwonipo for [G 0.5–1.0] (0.0)
  a1 says: puxedu
  a2 segments the context into 3 objects: 
       h-object-0 h-object-1 h-object-2
  a2 interprets puxedu as
       [G 0.5–1.0] (0.20)
  a2 identifies h-object-0 h-object-1
  a2 says: puxedu?
  a1 points to s-object-0
  a2 categorises the topic as [\textsc{width} 0.5–1.0]
  a2 stores puxedu as [\textsc{width} 0.5–1.0]
\end{verbatim}

@@

\begin{table}
\begin{center}
\begin{tabular}{lllll}
\lsptoprule
{\itshape Word}&{\itshape Meaning}&{\itshape Translation} & {\bfshape a1}&{\bfshape a2} \\ \midrule
wawosido & [\textsc{hpos} 0.0–0.5] &left&0.9&0.4\\ 
gixepo & [\textsc{hpos} 0.0–0.25] & very left&0.4&0.4\\ 
wonuxa & [\textsc{hpos} 0.0–0.25] & very left&0.10&0.0\\ 
meluri & [\textsc{hpos} 0.25–0.5] &medium left&0.4&0.4\\ 
gofubo & [\textsc{hpos} 0.5–1.0]& right&1.0&1.0\\ 
vokomutu & [\textsc{hpos} 0.5–0.75] &medium right&0.4&0.4\\ 
buwonipo & [\textsc{hpos} 0.75–0.875] &strongly right&0.1&0.0\\ 
wiwigapo & [\textsc{vpos} 0.0–0.5] &down&0.5&0.4\\ 
putuwenu & [\textsc{vpos} 0.5–1.0]&up & 0.5&0.5\\ 
vaviwumu & [\textsc{width} 0.5–1.0]&wide & 0.5&0.5\\ 
pesidumu & [\textsc{area} 0.0–0.5]&small& 0.2&0.2\\ 
fozumoba & [\textsc{area} 0.5–1.0]&large & 1.0&1.0\\ 
wefoto & [R 0.0–0.5]& low redness &0.2&0.1\\ 
togene & [R 0.5–1.0]& high redness &0.9&1.0\\ 
fumudanu & [G 0.0–0.5]& low greenness &0.7&0.9\\ 
puxedu & [G 0.5–1.0]& high greenness &0.9&0.9\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:puxedu}Population lexicon after 500 games.}
\end{center}
\end{table}
The global evolution of success and ontology is shown in the 
graph in \figref{psuccess1}. The graph is not 
fundamentally different from the ones we have
seen in the computer simulations in the previous
chapter, except that more failures occur after a lexicon
is established due to the contingencies of real 
world images and the unavoidable stochasticity 
associated with physical interactions.

@@

The RF-diagram\is{RF-diagram} in \figref{RF-diagram} shows the 
frequency with which certain words were used to communicate 
a particular situation. We see that in
the beginning the word `desepu' has been dominant, then 
there is a period of turbulence in which different 
words compete, but after a while a new word `va' wins 
the competition and becomes the dominant way to communicate 
about this situation. \figref{FM-diagram} shows another\is{FM-diagram}
competition diagram plotting the evolution of the FM-relation, for 
the word form `va', in other words the frequencies of 
all the meanings that co-occurred with the word `va'. 
We see an early peak when `va' was used in 70 \% of the 
games with the meaning [B 0.3125–0.375], i.e. a particular
shade of blue. Then there is a struggle during which additional
distinctions (on the \textsc{red} and \textsc{vpos}-channels) are competing
for the dominant meaning of `va'. 
[R 0.0–0.125] finally becomes dominant.

@@

These competition diagrams are an important tool to try 
and make sense of the ontological and lexical evolution
taking place in evolving groups of agents as they are
playing their language games. Typically we pick
a dominating word, for example the word `va' in 
\figref{RF-diagram}, and try to understand
why it has become dominant. 
The FM-diagram for `va' (\figref{FM-diagram}) explains
part of the story. Three stable meanings for `va' have emerged
around 1000 games: 
[R 0.0–0.125], [B 0.3125–0.375], and [\textsc{vpos} 0.25–0.5]. 
They are all equally adequate for distinguishing the object
`va' designates, and there are no situations yet
that would have forced the disambiguation of `va'. 
In game 3000, the environment (which is continuously 
changing in this experiment) produces a scene in which
a category which was distinctive for the object designated by 
`va' is no longer distinctive. The lexicon adapts 
immediately. Around game 3000 the \textsc{vpos}-based meaning disappeares, 
and the distinction based on \textsc{red} shoots up and becoming dominant.

@@

Different conceptualisations of the world (incoherence 
along the RM and MR dimensions) are even harder to 
avoid because every agent develops its own 
ontologies independently and without any direct 
feedback from the other agents. 
The ideal language system is not only impossible to 
attain for autonomous agents engaged in grounded
language games, it would be very 
inefficient to store and use as well because 
new words and meanings would be required for every new 
situation ever encountered. The larger the lexicon,
the harder the task of a language learner to acquire it. 
So there is a trade off between coherence and expressability. 
This is why natural languages constantly try to recruit existing 
words to keep down the repertoire of 
forms that have to be stored and hence learned.

@@

To conclude this chapter, I now discuss in more 
detail an example of this kind of semiotic dynamics as
gleaned from  an actual experiment with twenty Talking
Heads, taking turns to materialise themselves in two robotic 
bodies at a single physical site. The agents have 
only R,G,B, and gray-scale channels. This 
case study illustrates how the tools introduced in 
the previous section help us to make sense of
the very complex lexical evolution spontaneously 
arising in the system. We have set up the experiment
in such a way that the agents first see a limited set 
of objects. Then we
have progressively added new situations
to the environment (by pasting new figures on 
the white board or reconfiguring existing figures) and studied
the impact on the lexicons and ontologies of the agents.\footnote{
These experiments work were done in strong collaboration with 
Frederic Kaplan, see for example \cite{Steels:1999}.}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap7/figs/FM-FEPI.pdf}}
\caption{\label{fm-fepi}FM-diagram showing the different
meanings of `fepi'. One meaning [G 0.25–0.5] dominates.}
\end{figure}
We can see the meanings of `fepi', by 
inspecting its FM-diagram (\figref{fm-fepi}). 
The dominant meaning of `fepi' is 
a particular shade of green [G 0.25–0.5]. There are 
some other competing meanings (including [\textsc{gray} 0.25–0.3125])
but most of them are hardly ever used. We observe clearly 
the tendency for ambiguity to get damped.

What about synonymy? Let us look at the 
MF-diagram of [G 0.25–0.5], so that we can
see whether there are any other words in use for expressing
the same meaning (\figref{g02505.f}).
`fepi' has indeed emerged as dominant for this meaning, but 
this has not been without an intense struggle.
The tendency for synonymy to get damped is clearly present. 
Even though the lexicon occasionally destabilises, 
the lateral inhibition and the positive feedback loop between use 
and success causes self-organised MF-coherence. 
There is however something curious going on. 
In the early phases `xu' was dominant. Why did it 
destabilise and how has `fepi' has managed to become the
dominant expression for [G 0.25–0.5]?

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap7/figs/MF-G-025-050.pdf}}
\caption{\label{g02505.f}MF-diagram showing the
different words circulating in the population for expressing
the category [G 0.25–0.5]. First `xu' dominates and 
then `fepi' wins the competition after an intense struggle.}
\end{figure}

>>>>

But some drastic events happened in Wittgenstein's life which 
caused a radical shift in his philosophical orientation. 
One of them was that he became headmaster of a small
school in the Austrian mountains,
without losing his interest for philosophising about language and 
meaning. By working with children and by seeing concretely
how they engaged with language, Wittgenstein realised that the 
logicist approach did not address some basic questions one 
can ask about language, particularly how the semantics of a 
language might arise, or how people ever develop a shared 
communication system. Wittgenstein
saw clearly that a language imposes a certain view on the 
world, which gives each language its own strength as well as
its limitations. The logicist 
framework might be ideal as a tool for constructing a 
{\itshape post factum} formal description of the semantics of
an (ideal) language, just as a structuralist grammar identifies
the frozen state of its syntax, but
not for modeling how language achieves its communicative purpose, 
comes about, or evolves.

Wittgenstein proposed to study the use of language in terms of 
games. This notion captures that the most basic form of
language involves a social interaction between individuals
within a specific setting, which acts as a context restricting
the possible meanings. The interaction is played out 
along a conventionalised set of rules, as is chess or any
other game. The notion of a game captures
that a language interaction has a purpose, for example,
to identify an object in a context, to gather 
information, to transmit emotion, to try and
invoke an action by another person, etc. Words get \enlargethispage{2\baselineskip}
their meaning as part of a language game. Language is a tool 
fully integrated in the rest of human social activity.\footnote{See in particular \cite{Wittgenstein:1953}.}

@@

\begin{table}
\begin{center}
\begin{tabular}{lllll}
\lsptoprule
{\itshape Word}&{\itshape Meaning}&{\itshape Translation} & {\bfshape a1}&{\bfshape a2} \\ \midrule
wawosido & [\textsc{hpos} 0.0–0.5] &left&0.4&0.4\\ 
meluri & [\textsc{hpos} 0.25–0.5] &medium left&0.1&0.0\\ 
gofubo & [\textsc{hpos} 0.5–1.0]& right&1.0&1.0\\ 
wiwigapo & [\textsc{vpos} 0.0–0.5] &left&0.1&0.0\\ 
fozumoba & [\textsc{area} 0.5–1.0]&large & 0.0&0.1\\ 
wefoto & [\textsc{r} 0.0–0.5]& low redness &0.2&0.2\\ 
togene & [\textsc{r} 0.5–1.0]& high redness &0.5&1.0\\ 
fumudanu & [\textsc{g} 0.0–0.5]& low greenness &0.2&0.2\\ 
puxedu & [\textsc{g} 0.5–1.0]& high greenness &0.4&0.4\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game50}First words after about a dozen games.}
\end{center}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{lllll}
\lsptoprule
{\itshape Word}&{\itshape Meaning}&{\itshape Translation} & {\bfshape a1}&{\bfshape a2} \\ \midrule
wawosido & [\textsc{hpos} 0.0–0.5] &left&0.7&0.5\\ 
meluri & [\textsc{hpos} 0.25–0.5] &medium left&0.4&0.4\\ 
gofubo & [\textsc{hpos} 0.5–1.0]& right&1.0&1.0\\ 
vokomutu & [\textsc{hpos} 0.5–0.75] &medium right&0.2&0.2\\ 
buwonipo & [\textsc{hpos} 0.75–0.875] &strongly right&0.1&0.0\\ 
wiwigapo & [\textsc{vpos} 0.0–0.5] &down&0.6&0.6\\ 
fozumoba & [\textsc{area} 0.5–1.0]&large & 0.2&0.4\\ 
wefoto & [\textsc{r} 0.0–0.5]& low redness &0.2&0.3\\ 
togene & [\textsc{r} 0.5–1.0]& high redness &1.0&1.0\\ 
fumudanu & [\textsc{g} 0.0–0.5]& low greenness &0.7&0.7\\ 
puxedu & [\textsc{g} 0.5–1.0]& high greenness &0.7&0.9\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:goubo}Population lexicon after 100 games.}
\end{center}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{lll}
\lsptoprule
{\itshape channel}& {\itshape obj-0} & {\itshape obj-1}\\ \midrule
\textsc{hpos} & 0.08 & 0.09\\ 
\textsc{vpos} & 0.21 & 0.10\\ 
\textsc{height} & 0.10 & 0.15\\ 
\textsc{width} & 0.42 & 0.11\\ 
\textsc{area} & 0.34 & 0.16\\ 
\textsc{r} & 0.33 & 0.31\\ 
\textsc{g} & 0.68 & 0.65\\ 
\textsc{b} & 0.52 & 0.50\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:game126}Sensory data for game 126.}
\end{center}
\end{table}
Clearly width is the most salient channel for the speaker. 
It is therefore chosen and because the 
agents have already grown distinctions on this channel, 
discrimination succeeds and a new word can be
constructed and stored. 
\begin{verbatim}
Game 126 
  a2 is the speaker. a1 is the hearer. 
  a2 segments the context into 2 objects: 
       object-0 object-1
  a2 chooses object-0 as the topic 
  a2 categorises the topic as [\textsc{width} 0.5–1.0]
  a2 creates a new word: vaviwumu
  a2 says: vaviwumu
  a1 does not know vaviwumu
  a1 says: vaviwumu?
  a2 points to object-0
  a1 categorises the topic as [\textsc{width} 0.5–1.0]
  a1 stores vaviwumu as [\textsc{width} 0.5–1.0]
\end{verbatim}
\figref{fig:plate-13} (bottom) contains another example
of segmented images which have stimulated conceptual
growth and hence an expansion of the lexicon. 
In this case, categorisations on the \textsc{height} channel are
relevant and a word developed for tall.

@@

\begin{table}
\begin{center}
\begin{tabular}{lllll}
\lsptoprule
{\itshape Word}&{\itshape Meaning}&{\itshape Translation} & {\bfshape a1}&{\bfshape a2} \\ \midrule
wawosido & [\textsc{hpos} 0.0–0.5] &left&0.9&0.4\\ 
gixepo & [\textsc{hpos} 0.0–0.25] & very left&0.4&0.4\\ 
wonuxa & [\textsc{hpos} 0.0–0.25] & very left&0.10&0.0\\ 
meluri & [\textsc{hpos} 0.25–0.5] &medium left&0.4&0.4\\ 
gofubo & [\textsc{hpos} 0.5–1.0]& right&1.0&1.0\\ 
vokomutu & [\textsc{hpos} 0.5–0.75] &medium right&0.4&0.4\\ 
buwonipo & [\textsc{hpos} 0.75–0.875] &strongly right&0.1&0.0\\ 
wiwigapo & [\textsc{vpos} 0.0–0.5] &down&0.5&0.4\\ 
putuwenu & [\textsc{vpos} 0.5–1.0]&up & 0.5&0.5\\ 
vaviwumu & [\textsc{width} 0.5–1.0]&wide & 0.5&0.5\\ 
pesidumu & [\textsc{area} 0.0–0.5]&small& 0.2&0.2\\ 
fozumoba & [\textsc{area} 0.5–1.0]&large & 1.0&1.0\\ 
wefoto & [\textsc{r} 0.0–0.5]& low redness &0.2&0.1\\ 
togene & [\textsc{r} 0.5–1.0]& high redness &0.9&1.0\\ 
fumudanu & [\textsc{g} 0.0–0.5]& low greenness &0.7&0.9\\ 
puxedu & [\textsc{g} 0.5–1.0]& high greenness &0.9&0.9\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:upper}Population lexicon after 200 games.}
\end{center}
\end{table}
Note that new words have come into the lexicon 
for up (`putuwenu'), down (`wiwigapo'), and 
wide (`vaviwumu'). The discrimination trees of 
the \textsc{width} and \textsc{height} channel have started to 
expand (\figref{discri200} bottom).

@@

Grounding language games in physical environments\is{perceptual anomalies}
makes it obviously harder
for the agents in a number of respects. First of 
all because perception and segmentation may differ, 
the salient characteristics of objects may not be the same
and as a result the hearer may guess another meaning for an 
unknown word, compared to 
the one used by the speaker. This happens for
example in the following game (game 128) which is based on 
the segmented images shown in \figref{fig:plate-16} (top). The
speaker's perception is shown to the left and the 
hearer's to the right. 
\begin{verbatim}
Game 128
  a1 is the speaker. a2 is the hearer. 
  a1 segments the context into 3 objects: 
       object-0 object-1 object-2
  a1 chooses object-0 as the topic 
  a1 categorises the topic as 
       [HPOS 0.0–0.25] [HPOS 0.0–0.125]
  a1 creates a new word: `bagaxe' for [HPOS 0.0–0.25]
  a1 says: bagaxe
  a2 does not know bagaxe
  a2 says: bagaxe?
  a1 points to object-0
  a2 categorises the topic as [AREA 0.5–1.0]
  a2 stores bagaxe as [AREA 0.5–1.0]
\end{verbatim}

@@

Due to perceptual anomalies, a game may fail even though 
both agents already have the same meaning for the 
same word. This happens when this meaning yields different 
objects (or no objects at all) for the speaker and the hearer. 
As a consequence, the hearer adopts another meaning for 
the word which starts to compete with the one he already had. 
This happened in the following game (game 127)
based on the segmented images shown in \figref{fig:plate-16} (bottom). The speaker (left image) has identified
only two objects, but the hearer three. The 
speaker (left image) uses green as distinguishing 
characteristic which is indeed appropriate. but 
because the hearer's third object (the top rectangle) is 
also green, this distinction fails for him.  
Even though the hearer had already a well established
meaning for `puxedu', he associates a new meaning 
based on the \textsc{width}-channel with this word because this 
channel is now the most salient. 
\begin{verbatim}
Game 127
  a1 is the speaker. a2 is the hearer. 
  a1 segments the context into 2 objects: 
       s-object-0 s-object-1
  a1 chooses object-0 as the topic 
  a1 considers as salient G R AREA HPOS 
  a1 categorises the topic as 
       [HPOS 0.0–0.5] [HPOS 0.0–0.125]
       [AREA 0.0–0.5] [R 0.0–0.5] [G 0.5–1.0]
  a1 has the words
       puxedu for [G 0.5–1.0] (1.0)
       wawosido for [HPOS 0.0–0.5] (0.7)
       pesidimu for [AREA 0.0–0.5] (0.2)
       wefoto for [R 0.0–0.5] (0.2)
       buwonipo for [G 0.5–1.0] (0.0)
  a1 says: puxedu
  a2 segments the context into 3 objects: 
       h-object-0 h-object-1 h-object-2
  a2 interprets puxedu as
       [G 0.5–1.0] (0.20)
  a2 identifies h-object-0 h-object-1
  a2 says: puxedu?
  a1 points to s-object-0
  a2 categorises the topic as [WIDTH 0.5–1.0]
  a2 stores puxedu as [WIDTH 0.5–1.0]
\end{verbatim}

@@

\begin{table}
\begin{center}
\begin{tabular}{lllll}
\lsptoprule
{\itshape Word}&{\itshape Meaning}&{\itshape Translation} & {\bfshape a1}&{\bfshape a2} \\ \midrule
wawosido & [\textsc{hpos} 0.0–0.5] &left&0.9&0.4\\ 
gixepo & [\textsc{hpos} 0.0–0.25] & very left&0.4&0.4\\ 
wonuxa & [\textsc{hpos} 0.0–0.25] & very left&0.10&0.0\\ 
meluri & [\textsc{hpos} 0.25–0.5] &medium left&0.4&0.4\\ 
gofubo & [\textsc{hpos} 0.5–1.0]& right&1.0&1.0\\ 
vokomutu & [\textsc{hpos} 0.5–0.75] &medium right&0.4&0.4\\ 
buwonipo & [\textsc{hpos} 0.75–0.875] &strongly right&0.1&0.0\\ 
wiwigapo & [\textsc{vpos} 0.0–0.5] &down&0.5&0.4\\ 
putuwenu & [\textsc{vpos} 0.5–1.0]&up & 0.5&0.5\\ 
vaviwumu & [\textsc{width} 0.5–1.0]&wide & 0.5&0.5\\ 
pesidumu & [\textsc{area} 0.0–0.5]&small& 0.2&0.2\\ 
fozumoba & [\textsc{area} 0.5–1.0]&large & 1.0&1.0\\ 
wefoto & [\textsc{r} 0.0–0.5]& low redness &0.2&0.1\\ 
togene & [\textsc{r} 0.5–1.0]& high redness &0.9&1.0\\ 
fumudanu & [\textsc{g} 0.0–0.5]& low greenness &0.7&0.9\\ 
puxedu & [\textsc{g} 0.5–1.0]& high greenness &0.9&0.9\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:puxedu}Population lexicon after 500 games.}
\end{center}
\end{table}
The global evolution of success and ontology is shown in the 
graph in \figref{psuccess1}. The graph is not 
fundamentally different from the ones we have
seen in the computer simulations in the previous
chapter, except that more failures occur after a lexicon
is established due to the contingencies of real 
world images and the unavoidable stochasticity 
associated with physical interactions.

@@

The RF-diagram\is{RF-diagram} in \figref{RF-diagram} shows the 
frequency with which certain words were used to communicate 
a particular situation. We see that in
the beginning the word `desepu' has been dominant, then 
there is a period of turbulence in which different 
words compete, but after a while a new word `va' wins 
the competition and becomes the dominant way to communicate 
about this situation. \figref{FM-diagram} shows another\is{FM-diagram}
competition diagram plotting the evolution of the FM-relation, for 
the word form `va', in other words the frequencies of 
all the meanings that co-occurred with the word `va'. 
We see an early peak when `va' was used in 70 \% of the 
games with the meaning [\textsc{b} 0.3125–0.375], i.e. a particular
shade of blue. Then there is a struggle during which additional
distinctions (on the \textsc{red} and \textsc{vpos}-channels) are competing
for the dominant meaning of `va'. 
[\textsc{r} 0.0–0.125] finally becomes dominant.

@@

These competition diagrams are an important tool to try 
and make sense of the ontological and lexical evolution
taking place in evolving groups of agents as they are
playing their language games. Typically we pick
a dominating word, for example the word `va' in 
\figref{RF-diagram}, and try to understand
why it has become dominant. 
The FM-diagram for `va' (\figref{FM-diagram}) explains
part of the story. Three stable meanings for `va' have emerged
around 1000 games: 
[\textsc{r} 0.0–0.125], [\textsc{b} 0.3125–0.375], and [\textsc{vpos} 0.25–0.5]. 
They are all equally adequate for distinguishing the object
`va' designates, and there are no situations yet
that would have forced the disambiguation of `va'. 
In game 3000, the environment (which is continuously 
changing in this experiment) produces a scene in which
a category which was distinctive for the object designated by 
`va' is no longer distinctive. The lexicon adapts 
immediately. Around game 3000 the \textsc{vpos}-based meaning disappeares, 
and the distinction based on \textsc{red} shoots up and becoming dominant.

@@

Different conceptualisations of the world (incoherence 
along the RM and MR dimensions) are even harder to 
avoid because every agent develops its own 
ontologies independently and without any direct 
feedback from the other agents. 
The ideal language system is not only impossible to 
attain for autonomous agents engaged in grounded
language games, it would be very 
inefficient to store and use as well because 
new words and meanings would be required for every new 
situation ever encountered. The larger the lexicon,
the harder the task of a language learner to acquire it. 
So there is a trade off between coherence and expressability. 
This is why natural languages constantly try to recruit existing 
words to keep down the repertoire of \enlargethispage{2\baselineskip} 
forms that have to be stored and hence learned.

@@

To conclude this chapter, I now discuss in more 
detail an example of this kind of semiotic dynamics as
gleaned from  an actual experiment with twenty Talking
Heads, taking turns to materialise themselves in two robotic 
bodies at a single physical site. The agents have 
only R, G, B, and gray-scale channels. This 
case study illustrates how the tools introduced in 
the previous section help us to make sense of
the very complex lexical evolution spontaneously 
arising in the system. We have set up the experiment
in such a way that the agents first see a limited set 
of objects. Then we
have progressively added new situations
to the environment (by pasting new figures on 
the white board or reconfiguring existing figures) and studied
the impact on the lexicons and ontologies of the agents.\footnote{
These experiments work were done in strong collaboration with 
Frederic Kaplan, see for example \cite{Steels:1999}.}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap7/figs/FM-FEPI.pdf}}
\caption{\label{fm-fepi}FM-diagram showing the different
meanings of `fepi'. One meaning [\textsc{g} 0.25–0.5] dominates.}
\end{figure}
We can see the meanings of `fepi', by 
inspecting its FM-diagram (\figref{fm-fepi}). 
The dominant meaning of `fepi' is 
a particular shade of green [\textsc{g} 0.25–0.5]. There are 
some other competing meanings (including [\textsc{gray} 0.25–0.3125])
but most of them are hardly ever used. We observe clearly 
the tendency for ambiguity to get damped.

What about synonymy? Let us look at the 
MF-diagram of [\textsc{g} 0.25–0.5], so that we can
see whether there are any other words in use for expressing
the same meaning (\figref{g02505.f}).
`fepi' has indeed emerged as dominant for this meaning, but 
this has not been without an intense struggle.
The tendency for synonymy to get damped is clearly present. 
Even though the lexicon occasionally destabilises, 
the lateral inhibition and the positive feedback loop between use 
and success causes self-organised MF-coherence. 
There is however something curious going on. 
In the early phases `xu' was dominant. Why did it 
destabilise and how has `fepi' has managed to become the
dominant expression for [\textsc{g} 0.25–0.5]?

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap7/figs/MF-G-025-050.pdf}}
\caption{\label{g02505.f}MF-diagram showing the
different words circulating in the population for expressing
the category [\textsc{g} 0.25–0.5]. First `xu' dominates and 
then `fepi' wins the competition after an intense struggle.}
\end{figure}
